Accelerating adaptive ultrasound algorithms using graphics processing units \vspace{0.5cm}

### 2014-01-09 ThesisJP/Introduction/introduction.tex
Extended summary of Paper 1.

\section{Motivation}
A rapid development in computer game technology and accompanying programming languages have recently provided researchers with small personal super computers, comprised in a single graphics card (\nom{GPU}{Graphics Processing Unit}). The latest graphics cards\footnote{As of January 2014} from both Nvidia and AMD provides close to 6 T\nom{FLOPS}{Floating point operations per second} single precision and 1.5 TFLOPS double precision. For double precision this is more than the worlds most powerful supercomputer could provide in 1998.

@@

The Capon beamformer is a good example of an algorithm which has been highly studied for the last decade for the application of medical ultrasound imaging \todo{cite papers}, but where the number of calculations was though to be too much for real time processing. No wonder, to achieve real-time processing for e.g. cardiac ultrasound imaging an effective processing rate of around 100 MFLOPS is required. This number, as explained in Background \ref{sec:adaptbf}, also grows with larger subarrays, more image pixels, and higher framerate requirements.  With modern GPUs these levels of processing are available within a single card.

@@

My main focus have been the Capon beamformer, and the problem of making this computationally intense algorithm available for real-time ultrasound imaging. However, two additional methods have also been included in the thesis. The first is adaptive volume rendering of cardiac ultrasound volumes and the latter is simulation of dense ultrasound pressure fields. Both share the property of being computationally heavy, but they also consist of many independent computations which is perfect for parallel GPU computing.

Well on my way into the project, when a real-time Capon beamformer was realized, and loops of images for the first time were processed, new issues where discovered and had to be solved. This led me to a more theoretical study of the Capon beamformer in \textbf{Paper III}, with special attention on how to obtain high lateral resolution while preserving the important shift-invariant property of medical ultrasound imaging. Shift-invariant behavior is crucial if the method is going to be applied for live scanning.

@@

\subsection{Paper I}
\textbf{An Optimised GPU Implementation of the MVDR Beamformer for Active Sonar Imaging}\\
J.\:I.\:Buskenes, \textbf{J.\:P.\:\AA{}sen}, C.-I.\:C.\:Nilsen and A.\:Austeng\\
{\it IEEE Transactions on Oceanic Engineering, submitted.}\\\\
The first paper describes in details how the Capon beamformer is mapped to GPU architecture. Even though the paper is written within the field of active sonar imaging, the implementation described is also applicable to other active ultrasound imaging systems. A similar discussion for cardiac ultrasound imaging can be found in \textbf{Paper VI}.  Active sonar imaging typically differs from medical ultrasound imaging by a lower real-time requirement, element count, center frequency, and relative bandwidth.

The estimation of the spatial covariance matrix receives special attention. In previous literature on Capon beamforming the matrix inversion is always regarded as the most computationally complex piece.

>>>>

\section{Motivation}
A rapid development in computer game technology and accompanying programming languages have recently provided researchers with small personal super computers, comprised in a single graphics card (\nom{GPU}{Graphics Processing Unit}). The latest graphics cards\footnote{As of January 2014} from both \nom{Nvidia}{Graphics card producer} and \nom{AMD}{Graphics card producer} provides close to 6 T\nom{FLOPS}{Floating point operations per second} single precision and 1.5 TFLOPS double precision. For double precision this is more than the worlds most powerful supercomputer could provide in 1998.

@@

The Capon beamformer is a good example of an algorithm which has been highly studied for the last decade for the application of medical ultrasound imaging \todo{cite papers}, but where the number of calculations was though to be too much for real time processing. No wonder, to achieve real-time processing for e.g. cardiac ultrasound imaging an effective processing rate of around 100 MFLOPS is required. This number, as explained in Background \ref{sec:adaptbf}, also grows with larger subarrays, more image pixels, and higher framerate requirements.  With modern GPUs these levels of processing are finally available within a single card.

@@

My main focus have been the Capon beamformer, and the problem of making this computationally intense algorithm available for real-time ultrasound imaging (\textbf{Paper\,I} and \textbf{II}). However, two additional methods were also explored. The first is adaptive volume rendering of cardiac ultrasound volumes (\textbf{Paper\,IV}) and the latter is simulation of dense ultrasound pressure fields (\textbf{Paper\,V}). Both share the property of being computationally heavy, but they also consist of many independent computations which is perfect for parallel GPU computing.

Well on my way into the project, when a real-time Capon beamformer was realized, and loops of images for the first time were processed, new issues where discovered and had to be solved. This led me to a more theoretical study of the Capon beamformer (\textbf{Paper III}), with special attention on how to obtain high lateral resolution while preserving the important shift-invariant property of medical ultrasound imaging. Shift-invariant behavior is crucial if the method is ever to be applied for live scanning.

@@

\subsection{Paper I}
\textbf{An Optimised GPU Implementation of the MVDR Beamformer for Active Sonar Imaging}\\
J.\:I.\:Buskenes, \textbf{J.\:P.\:\AA{}sen}, C.-I.\:C.\:Nilsen and A.\:Austeng\\
{\it IEEE Transactions on Oceanic Engineering, submitted.}\\\\
The first paper describes in details how the Capon beamformer is mapped to GPU architecture. Even though the paper is written within the field of active sonar imaging, the implementation described is applicable to a vast range of active imaging systems. A similar discussion for cardiac ultrasound imaging can be found in \textbf{Paper VI}.  Active sonar imaging typically differs from medical ultrasound imaging by a lower real-time requirement, element count, center frequency, and relative bandwidth.

The estimation of the spatial covariance matrix receives special attention. In previous literature on Capon beamforming the matrix inversion has always been regarded as the most computationally complex piece. In this paper we show that estimation of the sample covariance matrix is actually the most complex part when spatial and temporal smoothing is applied with common parameters.
